---
title: "Web scraping to get the edge in your NBA fantasy draft"
output: html_document
date: "2023-10-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The website I will be getting the data from is the basketball reference website. I'm most interested in the game logs, as with these we can estimate each player's consistency throughout last season, and calculate fantasy scores.

```{r libraries, echo=FALSE}

library(tidyverse)
library(dplyr)
library(rvest)
library(XML)
library(glue)
library(janitor)
library(plotly)
library(RColorBrewer)

```

Investigating the pages containing each player game logs, we can see that basketball refernce uses a player code to construct the URL containing individual stats for each player. Using this URL as a template I have made the below function that can scrape the gamelog table for a given player code and year. 

This function utilises the rvest package function ```html_elements``` to parse the html code and find the table elements, then ```html_table``` reads this into a table format, which can be converted to a data frame. 

```{r get-player-codes}

# scrape game log table for a player in a given year
get_gamelog <- function(player_code, year){
  
  url <- paste0(glue('https://www.basketball-reference.com/players/i/{player_code}/gamelog/{year}'))
  # Scrape Player game log
  gamelog <- url %>%
    read_html() %>% 
    html_elements('table') %>%
    .[8] %>% 
    html_table() %>%
    as.data.frame()
  
  return(gamelog)

}

get_gamelog("jokicni01", 2023) %>% head()

```

However, I would ideally like to scrape the game logs for all players, or at least the top players from the 2023 season. To do this, we need to get all of the player codes. Thankfully, I was able to find a github repo with code that could be slightly modified to get a table from Basketball Reference containing all the player codes. 

I also decided to filter this player list, removing players with average minutes played below 25. 

```{r get-player-codes}

# TODO: Refactor this
# TODO: acknowledge references 

# Create Function to Scrape Player Links
scrape_nba_main <- function(yr){
  # Create URL
  url <- paste0('https://www.basketball-reference.com/leagues/NBA_', yr, '_per_game.html')
  webpage <- read_html(url)
  
  # Scrape All Player Links on Page
  webpage %>%
    html_nodes(xpath = "//td/a") %>% 
    html_attr("href") -> links
  links <- links[grepl("/players", links)]
  links <- links[!duplicated(links)]
  
  # Scrape Player Information - Age + Name
  webpage %>%
    html_nodes("table") %>%
    .[1] %>%
    html_table(fill = TRUE) %>%
    as.data.frame() -> yearx
  # print(yearx)
  yearx <- yearx %>% filter(Player != "Player")
  yearx <- yearx[!duplicated(yearx[c('Player', 'Age')]),]
  yearx <- yearx[,c(2,3,4,8)]
  # Fix Matt Thomas
  yearx <- yearx %>% filter(!(Player == 'Matt Thomas' & Age == ''))
  yearx$Link <- links
  yearx$Age <- NULL
  BBRefID <- strsplit(yearx$Link, '\\/')
  BBRefID <- sapply(BBRefID, function(x) x[4])
  BBRefID <- gsub(".html", "", BBRefID)
  yearx$BBRefID <- BBRefID
  # Return Data
  return(yearx)
}

#TODO: REFACTOR AGAIN 
# Run Code For 1950 to 2019
yearxs_nba <- 2023
all_data <- {}
for(i in 1:length(yearxs_nba)){
  data <- scrape_nba_main(yearxs_nba[i])
  all_data[i] <- list(data)
}

all_bb_ref <- do.call('rbind', all_data)
all_bb_ref <- all_bb_ref[!duplicated(all_bb_ref$BBRefID),]
all_bb_ref <- all_bb_ref[order(all_bb_ref$BBRefID),]

all_bb_ref_filtered <- all_bb_ref %>%
  mutate(MP = as.numeric(MP)) %>% 
  filter(MP > 25)

all_bb_ref_filtered %>%
  head()

```

OK, now that we have all of the basketball reference ID's its a matter of looping through them to construct the URL for each individual player's stats, scraping the game log table for each player and combining these into one master table. 

```{r scrape-all-gamelogs}

player_codes <- all_bb_ref$BBRefID
year <- 2023

all_game_logs <- list()
for (i in seq_along(player_codes)) {
  player_code <- player_codes[i]
  player_game_log <- get_gamelog(player_code=player_code, year=year) %>% 
    mutate(player_code = player_code) %>% 
    filter(Rk != "Rk") %>%  # remove the extra column-name rows 
    mutate(across(everything(), as.character)) 
    all_game_logs[[player_code]] <- player_game_log
  
  if (i > 1) {
    Sys.sleep(5) # sleep 5 seconds to prevent the website throttling the webscraper
    } 
}

all_game_logs_df <- bind_rows(all_game_logs)

all_game_logs_df %>%
  head()

```

Next, we need to clean the data. Rows corresponding to missed games have a number of different string values, so I replaced those with NA. Some rows that contained the column names also needed to be removed. I also added the player names into this Dataframe, using the player code to join it to the dataframe containing the player names.

I've also calculated number of double doubles, triple doubles, missed points, and a few other stats that are used to calculate fantasy scores, but not included in the original Basketball Reference table.

```{r, clean-data}

# these values correspond to when a player did not play in the game
missing_values <- c("Inactive|Did Not Dress|Did Not Play|Not With Team|Player Suspended")

# add in the player name
all_game_logs_df <- all_game_logs_df %>% 
  left_join(all_bb_ref[c("BBRefID", "Player")],
            by = c("player_code" = "BBRefID")) %>% 
  mutate_all(~str_replace_all(., missing_values, NA_character_)) %>% 
  mutate(across(c(
    "GS",
    "FG",
    "FGA",  
    "FG.",
    "X3P",
    "X3PA",
    "X3P.",
    "FT",
    "FTA",
    "FT.",
    "ORB",
    "DRB",
    "TRB", 
    "AST",
    "BLK",
    "TOV",
    "STL",
    "PTS"),
    ~as.numeric(.))) %>% 
  # make columns for triple doubles and double doubles:
  # DD: (IF: 2/5 OF ASS/BLOCK/STEAL/REB/PTS<9)
  # TD: (IF: 3/5 OF ASS/BLOCK/STEAL/REB/PTS<9)
  rowwise() %>% 
  mutate("DD" = if_else(sum(TRB > 9, AST > 9, BLK > 9, STL > 9, PTS > 9) >= 2, 1, 0)) %>%
  mutate("TD" = if_else(sum(TRB > 9, AST > 9, BLK > 9, STL > 9, PTS > 9) >= 3, 1, 0)) %>% 
  mutate("QD" = if_else(sum(TRB > 9, AST > 9, BLK > 9, STL > 9, PTS > 9) >= 4, 1, 0)) %>% 
  # games played - 1 if they played the game or 0 if not
  ungroup() %>% 
  mutate(GP = if_else(is.na(GS), 0, 1)) %>% # if game was played, 1 else zero
  # calculate number missed for field goals, free throws, three pointers
  mutate(FGM = FGA - FG) %>% 
  mutate(FTM = FTA - FT) %>% 
  mutate(X3PM = X3PA - X3P)

all_game_logs_df %>% 
  head()

```
Next, I decided to calculate fantasy scores for each player, to see how they would have performed in a fantasy league last year. I calculated total fantasy score, as well as mean, median and standard deviation. In order to gauge each players potential and consistency, 
These stats were calculated for the whole season (including missed games as zero's) to gauge each players consistency throughout the season - it's valuable to have players who actually play a lot of games and score highly throughout the season. To examine player potential, I also calculated the same stats with missed games removed - to see how well they play per-game. This metric should highlight any high-performing players who missed a lot of games last year. I also have added in player position and minutes played to this table.

Now we have a table which can definitively rank player performance in 2023, and will be very useful when making picks during the draft. 


```{r calculate-fantasy-scores}

# the sum of each metric * by a multiplier is used to calc the fantasy score 
# this is specific to the league 
multipliers_vector <- c(
  "GP" = 1,
  "FG" = 2,
  "FGM" = -1,
  "FT" = 1,
  "FTA" = 0.5,
  "FTM" = -1,
  "X3P" = 3.5,
  "X3PM" = -1.5,
  "ORB" = 3,
  "DRB" = 1,
  "TRB" = 1,
  "AST" = 4,
  "STL" = 5,
  "BLK" = 6,
  "TOV" = -2.5,
  "DD" = 10,
  "TD" = 30,
  "QD" = 1000,
  "PTS" = 1)

# calculate the fantasy score per game 
scores <- all_game_logs_df %>% 
  select(c(Player, names(multipliers_vector)))

weighted_scores = list()
for (i in seq_along(multipliers_vector)){
  name <- names(multipliers_vector[i])
  weighted_scores[[name]] <- scores[[name]] * multipliers_vector[name]
}
weighted_scores_df <- as.data.frame(weighted_scores)
agg_weighted_scores <- data.frame(
  player_name = scores$Player,
  weighted_scores = rowSums(weighted_scores_df, na.rm = T)
) %>%
  cbind(weighted_scores_df) %>% 
  mutate(weighted_scores_with_na = if_else(GP == 1, weighted_scores, NA)) # create another weighted score with NA's if they did not play, this will allow to calculate per-game stats 

final_weighted_scores_summarised <- agg_weighted_scores %>% 
  select(player_name, weighted_scores, weighted_scores_with_na, GP) %>% 
  group_by(player_name) %>%
  summarise(season_total_score = sum(weighted_scores),
            n_games_played = sum(GP), 
            season_mean = mean(weighted_scores), # season stats count missed games as zero
            season_median = median(weighted_scores),
            season_stdev = sd(weighted_scores), 
            games_played_mean = mean(weighted_scores_with_na, na.rm = TRUE), # games_played stats do not include missed games in the calculation
            games_played_median = median(weighted_scores_with_na, na.rm = TRUE),
            games_played_stdev = sd(weighted_scores_with_na, na.rm = TRUE)) %>%
  arrange(desc(season_median)) %>% 
  # add in other player stats from Basketball reference 
  left_join(all_bb_ref, by = c("player_name" = "Player")) %>% 
  select(-Link, -BBRefID)

final_weighted_scores_summarised %>%
  head()

```
```{r plots}

theme_set(theme_bw())

medians_scatter <- final_weighted_scores_summarised %>% 
  ggplot(aes(x = season_median,
             y = games_played_median,
             colour = Pos,
             text = glue(
               "
               Player: {player_name}
               Position: {Pos}
               Total score for season: {season_total_score}
               Median score for season: {season_median}
               Median score for games played: {games_played_median}
               "
             ))) +
  labs(x = 'Median score for season',
       y = 'Median score for games played')+
  geom_point() +
  scale_color_brewer(palette = 'Set3')

medians_scatter %>% ggplotly(tooltip = 'text')

```
```{r}

split_medians_scatter <- final_weighted_scores_summarised %>% 
  ggplot(aes(x = season_median,
             y = games_played_median,
             colour = Pos,
             text = glue(
               "
               Player: {player_name}
               Position: {Pos}
               Total score for season: {season_total_score}
               Median score for season: {season_median}
               Median score for games played: {games_played_median}
               "
             ))) +
  labs(x = 'Median score for season',
       y = 'Median score for games played')+
  geom_point() +
  facet_wrap(~Pos) +
  scale_color_brewer(palette = 'Set3')

split_medians_scatter %>% ggplotly(tooltip = 'text')

```

```{r}

total_scatter <- final_weighted_scores_summarised %>% 
  ggplot(aes(x = season_total_score,
             y = games_played_median,
             colour = Pos,
             text = glue(
               "
               Player: {player_name}
               Position: {Pos}
               Total score for season: {season_total_score}
               Median score for season: {season_median}
               Median score for games played: {games_played_median}
               "
             ))) +
  labs(x = 'Total score for season',
       y = 'Median score for games played')+
  geom_point() +
  scale_color_brewer(palette = 'Set3')

total_scatter %>% ggplotly(tooltip = 'text')

```

```{r}

split_total_scatter <- final_weighted_scores_summarised %>% 
  ggplot(aes(x = season_total_score,
             y = games_played_median,
             colour = Pos,
             text = glue(
               "
               Player: {player_name}
               Position: {Pos}
               Total score for season: {season_total_score}
               Median score for season: {season_median}
               Median score for games played: {games_played_median}
               "
             ))) +
  facet_wrap(~Pos) +
  labs(x = 'Total score for season',
       y = 'Median score for games played')+
  geom_point()+
  scale_color_brewer(palette = 'Set3')

split_total_scatter %>%
  ggplotly(tooltip = 'text')

```

```{r}
```


